---
title: "Building a database agent with ellmer"
output: rmarkdown::html_vignette
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Building a database agent with ellmer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

`oRm` ships an optional integration with the
[`ellmer`](https://ellmer.tidyverse.org) package that turns a `Chat` object into
a **database agent**. You hand `register_db_tools()` a chat plus one or more
models (or a whole `Engine`), and the chat is augmented with generic CRUD tools
the LLM can call to read, create, update, and delete rows.

This is an optional add-on: `ellmer` lives in `Suggests`, so `oRm` installs and
loads without it. `register_db_tools()` only requires `ellmer` when you actually
call it.

## Setup

```{r}
library(oRm)
library(ellmer)

engine <- Engine$new(
  drv = RSQLite::SQLite(),
  dbname = ":memory:",
  persist = TRUE
)

User <- engine$model(
  "users",
  id = Column("INTEGER", primary_key = TRUE, nullable = FALSE),
  name = Column("TEXT", nullable = FALSE),
  age = Column("INTEGER")
)
User$create_table()
```

## Registering the tools

Pass the chat and the model(s) you want to expose. The chat is modified in place
(ellmer chats are reference objects), so you can ignore the return value.

```{r}
chat <- chat_anthropic()
register_db_tools(chat, User)

chat$chat("Add a user named Kent who is 35, then list everyone older than 30.")
```

Four tools are registered:

- `db_read(table, filter)` — read rows; `filter` is an optional dplyr-style
  expression as a string, e.g. `"age > 30 & name == 'Kent'"`.
- `db_create(table, values)` — insert a row; `values` is a JSON object such as
  `{"name":"Kent","age":35}`.
- `db_update(table, filter, values)` — update matching rows (filter required).
- `db_delete(table, filter)` — delete matching rows (filter required).

The schema (every table and its columns) is embedded in each tool's description,
so the model can discover what's available without an extra call.

### Why dplyr filters?

Filter expressions are handed straight to `oRm`'s existing read/update/delete
machinery, which translates them to SQL via `dbplyr`. LLMs are fluent in dplyr,
so letting them write `age > 30 & name == 'Kent'` is both natural for the model
and reuses the same query builder you use elsewhere in `oRm`.

## Exposing several tables, or a whole schema

You can pass multiple models, and/or an `Engine` whose tables are reflected
automatically:

```{r}
Post <- engine$model(
  "posts",
  id = Column("INTEGER", primary_key = TRUE),
  user_id = ForeignKey("INTEGER", references = "users.id"),
  title = Column("TEXT")
)
Post$create_table()

# Explicit models ...
register_db_tools(chat, User, Post)

# ... or reflect everything the engine can see:
register_db_tools(chat, engine)

# Limit or skip tables when reflecting:
register_db_tools(chat, engine, tables = c("users", "posts"))
register_db_tools(chat, engine, exclude = "audit_log")
```

## Controlling write access

There is intentionally **no separate permission switch** in
`register_db_tools()`. Write access is governed by the `Engine`: open it
read-only and the create/update/delete tools fail with the engine's read-only
error, which is returned to the model so it can explain the limitation.

```{r}
ro_engine <- Engine$new(
  drv = RSQLite::SQLite(),
  dbname = "app.sqlite",
  .read_only = TRUE
)

chat <- chat_anthropic()
register_db_tools(chat, ro_engine)   # read works; writes are refused
```

## A note on trust

The agent runs SQL with the privileges of the connection you give it, and filter
expressions authored by the model are parsed and evaluated. `oRm` evaluates them
in a minimal environment so they cannot reach objects in your R session, but you
are still letting an LLM drive your database. For anything exposed to untrusted
input, pair the agent with a **read-only engine** and a connection scoped to only
the tables it should touch.
