---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

oRm is an object-relational mapping (ORM) framework designed for R users to work with SQL. Its core building blocks follow a chain of responsibility:

- `Engine`: Manages the database connection and model registry.
- `TableModel`: Represents a database table with column definitions and relationships.
- `Record`: Represents a row in a table. Supports CRUD operations.
- `Relationship`: Defines how models are linked, supporting joins and nested querying.

We’ll walk through each concept, starting with the `Engine`.

---

## Set up the Engine

```{r}
library(oRm)
```

```{r}
engine <- Engine$new(
  drv = RSQLite::SQLite(),
  dbname = ":memory:",
  persist = TRUE  # Optional for in-memory databases
)
```

### What the Engine Does

- Creates and manages the DBI connection.  
- Registers models so you can reference them by name or relationship.  
- Optionally uses pool for connection pooling (set use_pool = TRUE).  

You’ll rarely need to interact with the connection directly, but you can:

```{r, eval=FALSE}
engine$get_connection()
engine$list_tables()
engine$execute("SELECT * FROM users")
```

By default, connections are closed automatically after each operation unless persist = TRUE or use_pool = TRUE.

## Define a TableModel

TableModels can be created in two ways. The first is by calling the `TableModel` constructor directly:

```{r}
Users <- TableModel$new(
  "users", 
  engine, 
  id = Column("INTEGER", primary_key = TRUE),
  organization_id = ForeignKey("INTEGER", references = 'organizations.id'),
  name = Column("TEXT"),
  age = Column('INTEGER', default = 18)
)
```

Or, more commonly, you define a model through the engine itself:

```{r}
Organization <- engine$model(
  "organizations",
  id = Column("INTEGER", primary_key = TRUE),
  name = Column("TEXT")
)
```

This second approach automatically registers the model with the engine for use in relationships and queries.

## What TableModels Do

A `TableModel` instance gives you access to common operations on your table: 

### Create the table in your database

```{r}
Users$create_table()
Users$record(id = 1, name='John')$create()
Users$record(id = 2, name='Jane', age = 35)$create()
```

This creates the table based on your column definitions if it doesn’t already exist.

### Read rows from the table

```{r}
all_users <- Users$read()
young_users <- Users$read(age < 30)
```

The `read()` method accepts `dbplyr`-style filter conditions through `...`, allowing flexible querying using R expressions. It returns a list of `Record` objects, or a single record if `.mode = "get"` is specified.

```{r}
specific_user <- Users$read(id == 1, .mode = "get")
```

## What Records Do

Each row in a table is represented by a `Record`. Records provide methods for creating, updating, deleting, and accessing individual rows.

### Create a new record

```{r}
Users$record(id = 3, organization_id = 1, name = "Alice")$create()
```

### Update a record

```{r}
alice <- Users$read(id == 3, .mode = "get")
alice$data$name <- "Alicia"
alice$update()
```

### Delete a record

```{r}
alice$delete()
```

### Access record data

```{r}
print(alice$data$name)
```

## Defining and Using Relationships

You can define relationships between tables to enable seamless navigation between related records.

### Define a relationship

```{r}
define_relationship(
  Users,
  local_key = "organization_id",
  type = "many_to_one",
  related_model = Organization,
  related_key = "id",
  ref = "organization",
  backref = "users"
)
```

This allows records in `Users` to access their related `Organization`, and records in `Organization` to access all related `Users`.

### Accessing relationships through a Record

We defined the Organization earlier, but the table itself was nevver created. Let's create our table and give it an Organization to work with. 

```{r}
Organization$create_table()
Organization$record(id = 1, name = "Widgets, Inc")$create()
```

```{r}

Users$record(id = 3, name = 'Alice', organization_id = 1)$create()
alice = Users$read(id == 3, .mode='get')
alice_org <- alice$relationship('organization')
print(alice_org$data$name)
```

### Accessing relationships through a TableModel

```{r}
young_orgs <- Organization$relationship("users", age < 30)
young_orgs
```

This returns a list of user records with `age < 30` that belong to each organization.