---
title: "Hydrating Models from Existing Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydrating Models from Existing Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When a database already exists, `oRm` can introspect its tables and return
ready-to-use `TableModel` objects — no need to re-declare every column by hand.
This is called *hydration*.

## Single-table hydration with `hydrate()`

`engine$hydrate()` inspects one table and returns a `TableModel`:

```{r, eval=FALSE}
library(oRm)

engine <- Engine$new(
  drv    = RPostgres::Postgres(),
  dbname = "mydb",
  host   = "localhost"
)

# Reflect every column of the existing "users" table
Users <- engine$hydrate("users")
names(Users$fields)
#> [1] "id" "name" "email" "created_at"

# Narrow the model with include / exclude
Users <- engine$hydrate("users", include = c("id", "name", "email"))
Users <- engine$hydrate("users", exclude = c("password_hash", "internal_notes"))

Users$read(.mode = "data.frame")
```

### Reflection depth by dialect

| Dialect | What is reflected |
|---|---|
| Default (all) | Column names and best-effort types only |
| **PostgreSQL** | Canonical types, primary keys, nullability, column defaults, and foreign keys |

With PostgreSQL, `hydrate()` captures:

- **Canonical types** — e.g. `integer`, `text`, `timestamp with time zone`
- **Primary key flags** — `update()` and `delete()` work without any extra arguments
- **Nullability** — honoured on `create()`
- **Column defaults** — server-side defaults such as `nextval()` or `now()` are
  stored as `dbplyr::sql()` objects and applied by the database at insert time;
  `flush()` reads the computed value back into the record
- **Foreign keys** — reflected as `ForeignKey` objects, schema-qualified when
  the target lives in another schema

### Overriding reflected columns

Arguments passed via `...` take precedence over reflected columns, just like
`engine$model()`. This lets you attach a primary key on dialects that don't
reflect one, override a type, or add a `Method()`:

```{r, eval=FALSE}
# Supply the PK explicitly on a non-PostgreSQL backend
Users <- engine$hydrate(
  "users",
  id = Column("INTEGER", primary_key = TRUE)
)

# Attach a custom method alongside the reflected columns
Users <- engine$hydrate(
  "users",
  display_name = Method("table", function(self) {
    self$read(.mode = "data.frame")$name
  })
)
```

## Schema-wide hydration with `hydrate_schema()`

`engine$hydrate_schema()` hydrates every table in a schema at once and
automatically wires the `many_to_one` / `one_to_many` relationships implied by
the reflected foreign keys.

```{r, eval=FALSE}
# Hydrate all tables in the engine's default schema
models <- engine$hydrate_schema()

# Or restrict to a subset
models <- engine$hydrate_schema(tables = c("users", "posts", "comments"))

# Exclude tables you don't need
models <- engine$hydrate_schema(exclude = c("schema_migrations", "audit_log"))
```

### Accessing models

`hydrate_schema()` returns a named list keyed by bare table name:

```{r, eval=FALSE}
posts  <- models$posts
users  <- models$users
```

### Auto-wired relationships

Every reflected `ForeignKey` whose target was also hydrated is turned into a
`many_to_one` relationship (with the reverse `one_to_many` backref):

```{r, eval=FALSE}
post   <- posts$read(id == 1, .mode = "get")
author <- post$relationship("users")   # posts.user_id -> users.id
author$data$name
#> [1] "Kent"

all_posts <- author$relationship("posts")  # reverse backref
length(all_posts)
#> [1] 12
```

Foreign keys pointing at tables outside the hydrated set are skipped with a
warning. You can always wire those manually afterwards:

```{r, eval=FALSE}
models$posts$define_relationship(
  local_key     = "category_id",
  type          = "many_to_one",
  related_model = some_other_model,
  related_key   = "id",
  ref           = "category",
  backref       = "posts"
)
```

### Key arguments

| Argument | Default | Description |
|---|---|---|
| `tables` | `NULL` | Tables to hydrate; `NULL` = all in schema |
| `exclude` | `NULL` | Table names to skip |
| `.schema` | engine schema | Schema to inspect |
| `wire_relationships` | `TRUE` | Auto-wire FK-based relationships |

## Schema-qualified foreign keys

`ForeignKey()` supports cross-schema references in both shorthand and explicit
forms:

```{r, eval=FALSE}
# Shorthand: "schema.table.column"
fk <- ForeignKey("INTEGER", references = "audit.users.id")

# Explicit
fk <- ForeignKey(
  "INTEGER",
  ref_schema = "audit",
  ref_table  = "users",
  ref_column = "id",
  on_delete  = "CASCADE"
)
```

The generated SQL qualifies the target correctly:
`REFERENCES "audit"."users" ("id")`.

PostgreSQL reflection automatically schema-qualifies foreign keys when the
target lives in a different schema, so `hydrate_schema()` handles cross-schema
FK wiring transparently.
