---
title: "Using Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, include = FALSE}
library(oRm)

engine <- Engine$new(
    drv = RSQLite::SQLite(),
    dbname = ":memory:",
    persist = TRUE
)
```

Methods let you attach behavior to models and records, keeping business logic close to the data. This vignette starts with a minimal model and adds table- and record-level helpers.

### Methods

If database records are a model where columns are attributes, you may want reliable methods to alter, augment, or use those records. When creating a `TableModel`, use `Method()` to define methods for your model. Methods are declared at model creation, so each example builds a fresh model to keep it focused.

You can define methods for both the table level and the record level. Let's look at a table level method for students that performs a custom search:

```{r}
Student = engine$model(
    "students",
    id = Column('int'),
    name = Column('varchar'),
    age = Column('int'),

    search_by_name = Method(function(string) {
        self$read(dplyr::sql(paste0("name like '%", string, "%'")))
    }, target='table')
)
Student$create_table(overwrite = TRUE)

jane = Student$record(id=1, name = "Jane", age= 22)$create()
john = Student$record(id=2, name = "John", age = 25)$create()
tim  = Student$record(id=3, name = "Tim", age = 33)$create()

Student$search_by_name('j')

```

You can also have record level methods (this is the default behavior):


```{r}
Student = engine$model(
    "students",
    id = Column('int'),
    name = Column('varchar'),
    age = Column('int'),

    search_by_name = Method(function(string) {
        self$read(dplyr::sql(paste0("name like '%", string, "%'")))
    }, target='table'),

    greet = Method(function() {
        print(paste("Hi, my name is", self$data$name))
    })
)


Student$create_table(overwrite = TRUE)

jane = Student$record(id=1, name = "Jane", age= 22)$create()
jane$greet()

```

## Applications in Business Logic

Sometimes your model needs reusable business logic, not just CRUD. Methods are a good place to put that logic so it stays close to the data and is easy to test. We'll add a `Grades` table, then give `Students` a method to assign a grade and another to calculate an average.

First, define the grades table:

```{r}
Grades = engine$model(
    "grades",
    id = Column('int', primary_key = TRUE),
    class = Column('varchar'),
    student_id = ForeignKey('int', references = 'students.id'),
    value = Column('real')
)

Grades$create_table(overwrite=TRUE)
```

Now define a student model with three methods. `increment_id()` is a helper that calculates the next `Grades` id (SQLite does not auto-increment unless you declare it), `assign_grade()` creates a related grade record using that helper, and `get_average()` summarizes a student's grades.

```{r}
Students = engine$model(
    "students",
    id = Column('int', primary_key = TRUE),
    name = Column('varchar'),
    age = Column('int'),

    increment_id = Method(function() {
        last_grade = Grades$read(id == max(id), .mode = 'one_or_none')
        last_id = if (is.null(last_grade)) 0 else last_grade$data$id
        last_id + 1
    }),

    assign_grade = Method(function(class, value) {
        grade = Grades$record(
            id = self$increment_id(),
            class = class,
            student_id = self$data$id,
            value = value
        )$create()

        return(grade)
    }),

    get_average = Method(function() {
        grades = self$relationship('grades')
        mean(sapply(grades, \(x) x$data$value), na.rm=TRUE)
    })
)
Students$create_table(overwrite = TRUE)
```

With both tables in place, define the relationship so records can find their grades:

```{r}
define_relationship(
    Students, 'id', 'one_to_many', Grades, 'student_id',
    ref = 'grades',
    backref = 'student'
)
```

And now we can use the methods on a record:

```{r}
john = Students$record(id=1, name = "john", age = 22)$create()
john$assign_grade('Math', 83)
john$assign_grade('Math', 87)
john$relationship('grades')
john$get_average()
```
