---
title: "Using Engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

An `Engine` is a connection to a database. Under the hood, engines call `DBI::dbConnect()` 
to create and manage connections. This gives you flexibility to use any database supported 
by the `DBI` package. Connecting to a database with `Engine` is effectively the same as 
using `DBI::dbConnect()` directly.

So why wrap it in an R6 object?

- To manage opening and closing of connections
- To optionally support connection pooling
- To abstract some repetitive or database-specific boilerplate

### Creating an Engine

When creating an Engine, there are two additional arguments worth knowing:

- `persist = TRUE`: Keeps the connection open between operations like `model()` or `execute()`. 
  This is **required** when using an in-memory SQLite database, otherwise the data will be lost 
  between calls.
  
- `use_pool = TRUE`: Enables connection pooling (via the `pool` package), which improves performance 
  in environments like Shiny by reusing open connections across sessions.


```{r setup}
library(oRm)

engine <- Engine$new(
  drv = RSQLite::SQLite(),
  dbname = ":memory:",
  persist = TRUE
)

engine
```

### Common Methods

The full API is documented in the Engine reference page, but here are the most commonly used methods:

- `model()` Creates a `TableModel` object from the Engine. [See the TableModel vignette](./using-tablemodels.html)

- `get_connection()` Returns the underlying DBI connection. You can use this directly for raw SQL or with `dplyr::tbl()` for custom queries.

- `execute(sql)` A lightweight wrapper around `DBI::dbExecute()` using the Engine's connection.

- `get_query(sql)` Executes a query and returns the results as a data.frame using `DBI::dbGetQuery()`.  

```{r, eval=FALSE}

# Run a raw SQL statement
engine$execute("CREATE TABLE things (id INTEGER PRIMARY KEY, name TEXT)")

# Retrieve a result
df <- engine$get_query("SELECT * FROM things")


```

## Using with.Engine

`with.Engine` is an S3 method for managing transaction state. It allows you to run code 
within a database transaction, and if an error occurs during execution, it will 
automatically roll back the transaction.

This provides a safeguard against partial writes or inconsistent states. If you're familiar 
with Python's `sqlalchemy`, the usage will feel familiar.

```{r, eval=FALSE}
with(engine, {
  user = Users$record(name = "John Doe")
  user$create()
})
```

You can also choose to manually commit or roll back transactions. This gives you full
control over error handling:

```{r, eval=FALSE}
with(engine, {
  user = Users$record(name = "Jane Doe")
  tryCatch({
    user$create()
  }, error = function(e) {
    print(paste("An error occurred:", e$message))
    rollback()
  })
  commit()
}, auto_commit = FALSE)

```

This approach helps ensure that your data remains in sync with the database state.
If any part of the transaction fails, with.Engine() will automatically clean up
and alert you with a meaningful error.

## Dialects

Engines also store the dialect of the database they connect to. This is used to translate 
SQL statements into the correct syntax for the specific database backend.

Different databases handle data types, operators, and functions in slightly different ways. 
`oRm`'s dialect system isn't exhaustive, but it covers key differences that affect features 
like flushing or value-returning inserts.

For example:
- PostgreSQL supports `RETURNING *` to fetch inserted rows immediately.
- SQLite does not support `RETURNING`, so `oRm` falls back to using `last_insert_rowid()`.

These dialect-specific behaviors are handled automatically, so your code can stay consistent
regardless of the database in use.

## Read-Only Engines

Pass `.read_only = TRUE` when creating an Engine to prevent all write operations. This is
useful when giving analysts a connection to a production database, running exploratory
queries against live data, or any situation where accidental writes would be harmful.

```{r, eval=FALSE}
ro_engine <- Engine$new(
  drv    = RPostgres::Postgres(),
  dbname = "prod_db",
  host   = "db.example.com",
  .read_only = TRUE
)
```

### What is blocked

Any statement that is not a read (`SELECT`, `WITH`, `EXPLAIN`, `SHOW`, `PRAGMA`, `VALUES`)
is rejected with an error before it reaches the database:

```{r, eval=FALSE}
# Fine
ro_engine$get_query("SELECT count(*) FROM users")

# Blocked at the application level
ro_engine$execute("DELETE FROM users WHERE id = 1")
#> Error: Engine is read-only; refusing to execute non-SELECT statement.

# TableModel and Record write methods are also blocked
User$record(id = 99, name = "Ghost")$create()
#> Error: Engine is read-only; refusing write operation.

User$create_table()
#> Error: Engine is read-only; cannot create tables.
```

### Dialect-level enforcement

In addition to application-level guards, `oRm` applies a connection-level read-only flag
appropriate to each backend:

| Dialect    | Mechanism |
|------------|-----------|
| SQLite     | `RSQLite::SQLITE_RO` open flag — the connection itself cannot write |
| PostgreSQL | libpq `options="-c default_transaction_read_only=on"` — every transaction is read-only at the driver level |
| MySQL      | `SET SESSION TRANSACTION READ ONLY` executed after each new connection is opened |

For pooled MySQL connections, driver-level enforcement is not available; `oRm` will emit a
warning and fall back to application-level guards only.

